### 2021-07-05 rad-icfp-2021/intro.tex
US spelling "Generalization"

We propose a source-code transformation on a simple purely functional language for purely functional reverse-mode differentiation.
Our transformation is comprised of compilation scheme through a novel intermediate representation: unary form (UNF).
To emphasize the key ideas and for expository purposes, we present our work with a simple yet expressive language in Section~\ref{sec:simplediff}, 
and show how to extend our work to a richer language in Section~\ref{sec:generalisation}.

@@

Our contribution makes these claims precise. In more details, we introduce
\begin{itemize}
    \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\S\ref{sub:sourcelang})
    \item a method for extending our work and modular approaches to add and prove correct new primitives as needed in practice (\S\ref{sec:generalisation})
    \item a new intermediate representation (IR), UNF, essentially turning every operation into a unary one which inputs and outputs a tuple (\S\ref{sec:unf})
    \item a transformation between the IR and the main language (\S\ref{sub:transformations to and from UNF})
    \item a simple reverse-mode transformation on the IR (\S\ref{sub:Simple reverse mode transformation})
    \item the direct reverse-mode transformation obtained from the source to the target language (\S\ref{sub:Macro for pure reverse mode transformation})
    \item denotational semantics of our languages and transformations using multicategories and concategories (\S\ref{sec:correctness})
    \item correctness of the transformations (\S\ref{sec:correctness})
    \item complexity guarantees (\S\ref{sec:complexity})
    \item how efficiency is recovered via standard optimization techniques (\S\ref{sub:Optimizations})
    \item several additions to the source language (\S\ref{sec:generalisation}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
    \item a detailed comparison with other work and some limitations to our work (\S\ref{sec:conclusion})
\end{itemize}

@@

In Section~\ref{sec:background} we recall rudiments of automatic differentiation, forward and reverse-mode differentiation.
Next, in Section~\ref{sec:simplediff} we introduce the source and target languages for differentiation, and our reverse mode transformation. 
In Section~\ref{sec:unf} we introduce a new intermediate representation UNF, and a simple reverse-mode macro on this representation.
We introduce transformations from the source language to UNF and from UNF to the target language.
Next, in Section~\ref{sec:correctness} we give denotational semantics to our languages and show that our decomposition respects the diagram above and that our reverse mode transformation is correct.
In Section~\ref{sec:complexity}, we show that our transformation verifies the cheap gradient principle and extra general optimizations.
In Section~\ref{sec:generalisation} we show how to adapt our work for a richer source language. 
Finally, related work, limitations to our approach and future work are presented in Section~\ref{sec:conclusion}.

>>>>

We propose a source-code transformation on a simple purely functional language for purely functional reverse-mode differentiation.
Our transformation is comprised of a compilation scheme through a novel intermediate representation (IR): unary form (UNF).
To emphasize the key ideas and for expository purposes, we present our work with a simple yet expressive language in Section~\ref{sec:simplediff}, 
and show how to extend our work to a richer language in Section~\ref{sec:generalization}.

@@

Our contribution makes these claims precise. In more details, we introduce
\begin{itemize}
    \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\S\ref{sub:sourcelang})
    \item a method for extending our work and modular approaches to add and prove correct new primitives as needed in practice (\S\ref{sec:generalization})
    \item a new intermediate representation (IR), UNF, essentially turning every operation into a unary one which inputs and outputs a tuple (\S\ref{sec:unf})
    \item a transformation between the IR and the main language (\S\ref{sub:transformations to and from UNF})
    \item a simple reverse-mode transformation on the IR (\S\ref{sub:Simple reverse mode transformation})
    \item the direct reverse-mode transformation obtained from the source to the target language (\S\ref{sub:Macro for pure reverse mode transformation})
    \item denotational semantics of our languages and transformations using multicategories and concategories (\S\ref{sec:correctness})
    \item correctness of the transformations (\S\ref{sec:correctness})
    \item complexity guarantees (\S\ref{sec:complexity})
    \item how efficiency is recovered via standard optimization techniques (\S\ref{sub:Optimizations})
    \item several additions to the source language (\S\ref{sec:generalization}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
    \item a detailed comparison with other work and some limitations to our work (\S\ref{sec:conclusion})
\end{itemize}

@@

In Section~\ref{sec:background} we recall rudiments of automatic differentiation, forward and reverse-mode differentiation.
Next, in Section~\ref{sec:simplediff} we introduce the source and target languages for differentiation, and our reverse mode transformation. 
In Section~\ref{sec:unf} we introduce a new intermediate representation UNF, and a simple reverse-mode macro on this representation.
We introduce transformations from the source language to UNF and from UNF to the target language.
Next, in Section~\ref{sec:correctness} we give denotational semantics to our languages and show that our decomposition respects the diagram above and that our reverse mode transformation is correct.
In Section~\ref{sec:complexity}, we show that our transformation verifies the cheap gradient principle and extra general optimizations.
In Section~\ref{sec:generalization} we show how to adapt our work for a richer source language. 
Finally, related work, limitations to our approach and future work are presented in Section~\ref{sec:conclusion}.

### 2021-07-05 rad-icfp-2021/purerad.tex
writing a bit in correctness
Intro until section 1.4 + US spelling
US spelling "Generalization"

We consider a standard call-by-value language. 
It consists of a first-order functional language with arrays and a few typical second-order array operations. 
The types !T1,T2! and terms !e1,e2! are given in Figure~\ref{fig:source_grammar}.
We have included a minimal set of array operations for the sake of illustration,  it is not hard to add more.
See Section~\ref{sec:generalisation}.

@@

The typing rules are given in Figure~\ref{fig:source_typesystem}. 
For scalar operations, we assume given a set of operations, including $+$ and $*$. 
!op1! and !op2! denote respectively a unary and a binary operation on reals. 
These operations represent smooth total functions, 
but again this can be easily generalised (\S\ref{sec:generalisation}).  
Typical examples include !cos, exp, +, *!. 
We use infix notation for binary operators.

!reduce! is a fold left operator for which the function is assumed to be associative, 
and the provided initial value should be a unit of the binary operation.
It is a well-known parallel friendly construct. For the sake of simplicity in the presentation, the bound function in !reduce! 
is restricted to having no free variables. For the same reason, we currently restrict to arrays of reals.
We show how to lift these restrictions in Section~\ref{sec:generalisation}.

@@

In Figure~\ref{fig:direct_diff_macro} we present our direct transformation from the source language to the target language for pure reverse mode differentiation.
We explain in the next Section (\ref{sub:Partial evaluation and optimisation}) how to optimize it further and compute the gradient.
Admittedly, this transformation may be hard to read, and it is not straightforward to show its correctness directly. 
In Section~\ref{sec:unf} we decompose this transformation into three simpler steps via a novel intermediate representation (UNF).

@@

\subsection{Partial evaluation and optimisation} % (fold)
\label{sub:Partial evaluation and optimisation}

@@

\begin{tabular}{{r c l}}
    $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 A) (shift1(scanl * 1 A))!\\
    $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
    $\nabla_A$!dot(A,B)! &:=& !B! 
\end{tabular}
\end{example}

>>>>

We consider a standard call-by-value language. 
It consists of a first-order functional language with arrays and a few typical second-order array operations. 
The types !T1,T2! and terms !e1,e2! are given in Figure~\ref{fig:source_grammar}.
We have included a minimal set of array operations for the sake of illustration,  it is not hard to add more.
See Section~\ref{sec:generalization}.

@@

The typing rules are given in Figure~\ref{fig:source_typesystem}. 
For scalar operations, we assume given a set of operations, including $+$ and $*$. 
!op1! and !op2! denote respectively a unary and a binary operation on reals. 
These operations represent smooth total functions, 
but again this can be easily generalized (\S\ref{sec:generalization}).  
Typical examples include !cos, exp, +, *!. 
We use infix notation for binary operators.

!reduce! is a fold left operator for which the function is assumed to be associative, 
and the provided initial value should be a unit of the binary operation.
It is a well-known parallel friendly construct. For the sake of simplicity in the presentation, the bound function in !reduce! 
is restricted to having no free variables. For the same reason, we currently restrict to arrays of reals.
We show how to lift these restrictions in Section~\ref{sec:generalization}.

@@

In Figure~\ref{fig:direct_diff_macro} we present our direct transformation from the source language to the target language for pure reverse mode differentiation.
We explain in the next Section (\ref{sub:Partial evaluation and optimization}) how to optimize it further and compute the gradient.
Admittedly, this transformation may be hard to read, and it is not straightforward to show its correctness directly. 
In Section~\ref{sec:unf} we decompose this transformation into three simpler steps via a novel intermediate representation (UNF).

@@

\subsection{Partial evaluation and optimization} % (fold)
\label{sub:Partial evaluation and optimization}

@@

\begin{tabular}{{r c l}}
    $\nabla_A$!prod(A)! &=& !map2 * (scanr * 1 A) (shift1(scanl * 1 A))!\\
    $\nabla_A$!sum(A)! &=& !map (x -> 1) A!\\
    $\nabla_A$!dot(A,B)! &=& !B! 
\end{tabular}
\end{example}
